---
title: "medicalcoder vs icdcomorbid"
output:
 rmarkdown::html_vignette:
   toc: true
   number_sections: false
bibliography: ../references.bib
vignette: >
 %\VignetteIndexEntry{medicalcoder vs icdcomorbid}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r, label = "setup", include = FALSE}
# IMPORTANT SYNTAX NOTE:
# Vignettes cannot use the pipeOp `|>` so the package can depend on R >= 3.5.0
# The _articles_ (website only) can use the pipeOp since these are not part of the source package.

library(kableExtra)
options(qwraps2_markup = "markdown")
options(knitr.kable.NA = '')
knitr::opts_chunk$set(collapse = TRUE, fig.align = "center")
```

# Introduction

The purpose of this article is to compare the API and results between
medicalcoder and the R package
[icdcomorbid](https://cran.r-project.org/package=icdcomorbid)
[@r-icdcomorbid].

```{r}
library(medicalcoder)
library(icdcomorbid)
packageVersion("icdcomorbid")
packageDescription("icdcomorbid")$Title
cat(packageDescription("icdcomorbid")$Description)
```

icdcomorbid provides [@quan2005] Charlson and Elixhauser comorbidities.  A
nice feature of icdcomorbid is the ability to use custom mappings between
ICD codes and comorbidities.

# Prepare Data for icdcomorbid
The example data set `mdcr` within medicalcoder is in a format that is ideal
for `medicalcoder::comorbidities()`.  To prepare that data set for use in the
`icdcomorbid::icd9_to_comorbid()`  and `icdcomorbid::icd10_to_comorbid()`,
the data needs to be in a "wide" format instead of the provided long format.

We start by splitting the data into ICD-9 and ICD-10 sets and add rows for any
missing patids.

```{r}
mdcr_icd9dx  <- subset(mdcr, icdv ==  9L & dx == 1L)
mdcr_icd10dx <- subset(mdcr, icdv == 10L & dx == 1L)
mdcr_icd9dx <-
  rbind(
    mdcr_icd9dx,
    data.frame(patid = setdiff(mdcr$patid, mdcr_icd9dx$patid), icdv = 9L, code = "", dx = 1L)
  )
mdcr_icd10dx <-
  rbind(
    mdcr_icd10dx,
    data.frame(patid = setdiff(mdcr$patid, mdcr_icd10dx$patid), icdv = 10L, code = "", dx = 1L)
  )
```

`icdcomorbid::long_to_wide()` is provided to transform data sets, but it is very
slow.  I waited about fifteen seconds for the following and then stopped the
process because there are more efficient ways.

```{r}
#  mdcr_icd9dx_wide <-
#    icdcomorbid::long_to_wide(
#      df = mdcr_icd9dx,
#      idx = "patid",
#      icd_cols = "code"
#    )
#
# killed process after 15 seconds
```

```{r}
tic <- Sys.time()

library(data.table)
setDT(mdcr_icd9dx)
setDT(mdcr_icd10dx)
mdcr_icd9dx[, DX := paste0("DX", seq_along(code)), by = .(patid)]
mdcr_icd10dx[, DX := paste0("DX", seq_along(code)), by = .(patid)]

mdcr_icd9dx_wide <-
  data.table::dcast(mdcr_icd9dx, patid ~ DX, value.var = "code")

mdcr_icd10dx_wide <-
  data.table::dcast(mdcr_icd10dx, patid ~ DX, value.var = "code")

toc <- Sys.time()
difftime(toc, tic, units = "secs")
```

# Charlson Comorbidities

A detailed comparison between `medicalcoder::comorbidities()` and the methods in
the icdcomorbid package will not be presented here. Note the time required to
apply the Charlson comorbidities to the `mdcr` data set via
`medicalcoder::comorbidities()`.

```{r}
tic <- Sys.time()

medicalcoder_charlson_results <-
  medicalcoder::comorbidities(
    data = mdcr,
    id.vars = "patid",
    icd.codes = "code",
    dx.var = "dx",
    icdv.var = "icdv",
    poa = 1L, # assume all codes are present on admission
    primarydx = 0L, # assume all codes are secondary diagnoses
    method = "charlson_quan2011"
  )

toc <- Sys.time()

difftime(toc, tic, units = "secs")
```

Compare that to the amount of time required to process just the first 1,000
patids via `icdcomorbid::icd9_to_comorbid()`.

```{r}
tic <- Sys.time()

icd9_to_comorbid_results <-
  icdcomorbid::icd9_to_comorbid(
    df = mdcr_icd9dx_wide[1:1000, ],
    idx = "patid",
    icd_cols = grep("^DX", names(mdcr_icd9dx_wide), value = TRUE),
    mapping = "charlson9"
  )

toc <- Sys.time()

difftime(toc, tic, units = "secs")
```

There are `r length(unique(mdcr$patid))` patids to assess.  To compare the
results between `medicalcoder::comorbidities()` and the results from the R
package icdcomorbid would require calling `icdcomorbid::icd9_to_comorbid()`,
`icd10_to_comorbid()`, aggregating the results, and scoring.

Based just on the time required to compute the results, medicalcoder is
superior to icdcomorbid.

# References

<!-- ----------------------------------------------------------------------- -->
<!--                              End of File                                -->
<!-- ----------------------------------------------------------------------- -->
