---
title: "medicalcoder vs multimorbidity"
output:
 rmarkdown::html_vignette:
   toc: true
   number_sections: false
bibliography: ../references.bib
vignette: >
 %\VignetteIndexEntry{medicalcoder vs multimorbidity}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r, label = "setup", include = FALSE}
# IMPORTANT SYNTAX NOTE:
# Vignettes cannot use the pipeOp `|>` so the package can depend on R >= 3.5.0
# The _articles_ (website only) can use the pipeOp since these are not part of the source package.

library(kableExtra)
options(qwraps2_markup = "markdown")
options(knitr.kable.NA = '')
knitr::opts_chunk$set(collapse = TRUE, fig.align = "center")
```

# Introduction

The purpose of this article is to compare the API and results between
medicalcoder and the R package
[multimorbidity](https://cran.r-project.org/package=multimorbidity)
[@r-multimorbidity].

```{r}
library(data.table)
library(medicalcoder)
library(multimorbidity)
packageVersion("multimorbidity")
packageDescription("multimorbidity")$Title
cat(packageDescription("multimorbidity")$Description)
```

# Prepare Data for multimorbidity
The example data set `mdcr` within medicalcoder is in a format that is ideal
for `medicalcoder::comorbidities()` and for the `multimorbidity::charlson()`
call.  The only issue that needs to be addressed is that
`multimorbidity::charlson()` is expecting to only see diagnostic codes so we
will omit the procedure codes from `mdcr`.

```{r}
mdcr_dx <- subset(mdcr, dx == 1)

# add any patid omit
mdcr_dx <-
  rbind(
    mdcr_dx,
    data.frame(
      patid = setdiff(mdcr$patid, mdcr_dx$patid),
      icdv = 10L, code = "", dx = 1L
    )
  )
```

# Charlson Comorbidities

A detailed comparison between `medicalcoder::comorbidities()` and the methods in
the `multimorbidity::charlson()` call is presented here.

```{r}
tic <- Sys.time()

medicalcoder_charlson_results <-
  medicalcoder::comorbidities(
    data = mdcr,
    id.vars = "patid",
    icd.codes = "code",
    dx.var = "dx",
    icdv.var = "icdv",
    poa = 1L, # assume all codes are present on admission
    primarydx = 0L, # assume all codes are secondary diagnoses
    method = "charlson_quan2011"
  )

toc <- Sys.time()

difftime(toc, tic, units = "secs")
```

Calling `multimorbidity::charlson()`
```{r}
tic <- Sys.time()

multimorbidity_charlson_results <-
  multimorbidity::charlson(
    dat = mdcr_dx,
    id  = patid,
    version = 19, # default - both ICD-9 and ICD-10 data
    version_var = icdv
  )

toc <- Sys.time()

difftime(toc, tic, units = "secs")
```

## Differences in the results

We build a `data.frame` `deltas` to explore the differences in the results between
`medicalcoder::comorbidities()` and `multimorbidity::charlson()`.  One important
difference is that `medicalcoder::comorbidities()` returns a `data.frame` with
the id variable names as passed to the function whereas
`multimorbidity::charlson()` returns a `data.frame` with the id column named id.

Also, there is a difference in the storage mode for the indicator columns.
`multimorbidity::charlson()` returns numeric columns whereas
`medicalcoder::comorbidities()` return integer columns.

```{r}
multimorbidity_charlson_results <-
  multimorbidity_charlson_results |>
  dplyr::mutate_if(is.numeric, as.integer)

deltas <-
  merge(
    x = medicalcoder_charlson_results,
    y = multimorbidity_charlson_results,
    all = TRUE,
    by.x = "patid",
    by.y = "id"
  )
setDT(deltas)
```

We will compare the columns between the two methods.  None of the results are
the same.
```{r}
dcolumns <- data.table::fread(text = "
medicalcoder | multimorbidity
aidshiv | charlson_hiv
mal | charlson_malig
cebvd |charlson_cerebro
copd | charlson_chronic_pulm
chf | charlson_chf
dem | charlson_dementia
dmc | charlson_diab_c
dm | charlson_diab_uc
hp | charlson_hemi_para
mld | charlson_mild_liv
msld | charlson_mod_sev_liv
mst | charlson_met_solid
mi | charlson_myocar
pud | charlson_peptic_ulcer
pvd | charlson_periph_vasc
rnd | charlson_renal
rhd | charlson_rheum")

Map(
  f = function (x, y) {
    z <- identical(deltas[[x]], deltas[[y]])
    z
  },
  x = dcolumns$medicalcoder,
  y = dcolumns$multimorbidity
) |>
do.call(c, args = _)
```

```{r, include = FALSE}
# expect all identical() to return FALSE
Map(
  f = function (x, y) {
    z <- identical(deltas[[x]], deltas[[y]])
    stopifnot(!z)
  },
  x = dcolumns$medicalcoder,
  y = dcolumns$multimorbidity
) |>
do.call(c, args = _)
```

Let's dig into one of the conditions: Rheumatic disease ("rhd" in
`medicalcoder::comorbidities()`, "charlson_rheum" in
`multimorbidity::charlson()`).

```{r}
deltas[rhd != charlson_rheum, .(patid, rhd, charlson_rheum)]
deltas[rhd != charlson_rheum, .N, keyby = .(rhd, charlson_rheum)]
```
```{r, include = FALSE}
stopifnot(
  deltas[rhd != charlson_rheum, .N, keyby = .(rhd, charlson_rheum)] == c(1, 0, 136)
)
```

All the differences appear to be from `medicalcoder::comorbidities()` flagging
the condition and `multimorbidity::charlson()` not flagging the condition.
Let's look at the ICD codes `medicalcoder::comorbidities()` flagged in the
records

```{r}
missingcodes <-
  merge(
    x = subset(mdcr, patid %in% deltas[rhd != charlson_rheum, patid]),
    y = subset(medicalcoder::get_charlson_codes(), charlson_quan2005 == 1L),
    by = c("code", "icdv", "dx")
  )
missingcodes <- subset(missingcodes, condition == "rhd")
missingcodes <- unique(missingcodes[, c("icdv", "dx", "code", "full_code")])
setDT(missingcodes)
missingcodes[, codeid := 1:.N]

missingcodes
```
As we see below, all these codes are flagged by `medicalcoder::comorbidities()`
and not flagged by `multimorbidity::charlson()`.
```{r}
medicalcoder::comorbidities(
  data = missingcodes,
  id.vars = "codeid",
  icd.codes = "code",
  icdv.var = "icdv",
  dx.var = "dx",
  method = "charlson_quan2005",
  poa = 1L,
  primarydx = 0L
)$rhd

multimorbidity::charlson(
  dat = missingcodes,
  id = codeid,
  version = 19,
  version_var = icdv
)$charlson_rheum
```

Referencing Table 1 of @quan2005, we see that the ICD-10 codes for Rheumatic
disease are:

> M05.x, M06.x, M31.5, M32.x–M34.x, M35.1, M35.3, M36.0

And the enhanced ICD-9-CM codes are:

> 446.5, 710.0–710.4, 714.0–714.2, 714.8, 725.x


# Conclusions

`multimorbidity::charlson()` takes more time to evaluate the same data set than
`medicalcoder::comorbidities()` and `multimorbidity::charlson()` reports several
false negatives.

`medicalcoder::comorbidities()` is a preferable utility.

# References

<!-- ----------------------------------------------------------------------- -->
<!--                              End of File                                -->
<!-- ----------------------------------------------------------------------- -->
