---
title: "medicalcoder vs mimic"
output:
 rmarkdown::html_vignette:
   toc: true
   number_sections: false
bibliography: ../references.bib
vignette: >
 %\VignetteIndexEntry{medicalcoder vs mimic}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r, label = "setup", include = FALSE}
# IMPORTANT SYNTAX NOTE:
# Vignettes cannot use the pipeOp `|>` so the package can depend on R >= 3.5.0
# The _articles_ (website only) can use the pipeOp since these are not part of the source package.

library(kableExtra)
options(qwraps2_markup = "markdown")
options(knitr.kable.NA = '')
knitr::opts_chunk$set(collapse = TRUE, fig.align = "center")
```

# Introduction

The purpose of this article is to compare the results for the Charlson comorbidity
flags returned by `medicalcoder::comorbidities()` vs the SQL code provided by
@johnson2018mimic.

```{r}
library(medicalcoder)
```

# MIMIC SQL

SQL code for applying a Charlson comorbidity algorithm to the MIMIC-IV data is
available from MIT Laboratory for Computational Physiology (MIT_LCP) on
GitHub\footnote{\url{https://github.com/MIT-LCP/mimic-code.git}}.  The code is
expected to run on Google Big Query on the MIMIC-IV data.  We make a few small
modifications to the code so we can evaluate the SQL locally via RSQLite and on
a local data set.

```{r}
mimic_charson_query <-
  scan(
    file = "https://raw.githubusercontent.com/MIT-LCP/mimic-code/278df75ec30991ff3a6f5ceb6d2221635a085e9f/mimic-iv/concepts/comorbidity/charlson.sql",
    what = character(),
    sep = "\n"
  )

# modify the query to work in SQLite
mimic_charson_query <-
  gsub(pattern = "physionet-data.mimiciv_hosp.admissions",
       replacement = "admissions",
       x = mimic_charson_query,
       fixed = TRUE)
mimic_charson_query <-
  gsub(pattern = "physionet-data.mimiciv_hosp.diagnoses_icd",
       replacement = "diagnoses",
       x = mimic_charson_query,
       fixed = TRUE)
mimic_charson_query <-
  gsub(pattern = "physionet-data.mimiciv_derived.age",
       replacement = "ages",
       x = mimic_charson_query,
       fixed = TRUE)
mimic_charson_query <-
  gsub(pattern = "GREATEST",
       replacement = "MAX",
       x = mimic_charson_query,
       fixed = TRUE)

mimic_charson_query <- paste(mimic_charson_query, collapse = "\n")
```

```{r}
library(medicalcoder)
library(data.table)
mdcr_for_mimic <- data.table::copy(mdcr)
setDT(mdcr_for_mimic)

# create a hospital admission id  (one admission per patient id in mdcr)
setnames(
  mdcr_for_mimic,
  old = c("patid", "code", "icdv"),
  new = c("subject_id", "icd_code", "icd_version")
)

mdcr_for_mimic[, hadm_id := paste0(subject_id, "e1")]
mdcr_for_mimic[, seq_num := 1L:.N, by = .(subject_id, hadm_id)]
mdcr_for_mimic[, age := as.integer(substr(as.character(subject_id), 1L, 2L))]

```

```{r}
library(odbc)
library(DBI)
library(RSQLite)

con <- dbConnect(drv = RSQLite::SQLite(), dbname = ":memory:")

# add data to the data base
dbWriteTable(conn = con, name = "diagnoses",  value = mdcr_for_mimic[dx == 1L])
dbWriteTable(conn = con, name = "admissions", value = mdcr_for_mimic[, unique(.SD), .SDcols = c("subject_id", "hadm_id")])
dbWriteTable(conn = con, name = "ages",       value = mdcr_for_mimic[, unique(.SD), .SDcols = c("hadm_id", "age")])

# get the charlson results via MIMIC-IV
mimic_charlson_results <- dbGetQuery(con, mimic_charson_query)

# close DB connection
dbDisconnect(conn = con)

setDT(mimic_charlson_results)
```

```{r}
medicalcoder_charlson_results <-
  comorbidities(
    data = mdcr_for_mimic,
    id.vars = c("subject_id", "hadm_id"),
    icd.codes = "icd_code",
    icdv.var = "icd_version",
    dx.var = "dx",
    age.var = "age",
    method = "charlson_quan2005",
    full.codes = FALSE,
    flag.method = "current",
    poa = 1L,
    primarydx = 0L
  )
```

```{r}
delta <-
  merge(
    x = medicalcoder_charlson_results,
    y = mimic_charlson_results,
    all = TRUE,
    by = c("subject_id", "hadm_id")
  )
```


```{r}
uniqueN(mdcr_for_mimic$hadm_id)
nrow(mimic_charlson_results)
nrow(medicalcoder_charlson_results)
```

Conditions without multiple severity levels are the same between the two
methods.
```{r}
dcolumns <- fread(text = "
medicalcoder | mimic
aidshiv      | aids
cebvd        | cerebrovascular_disease
copd         | chronic_pulmonary_disease
chf          | congestive_heart_failure
dem          | dementia
hp           | paraplegia
mi           | myocardial_infarct
pud          | peptic_ulcer_disease
pvd          | peripheral_vascular_disease
rnd          | renal_disease
rhd          | rheumatic_disease
age_score.x  | age_score.y
cci          | charlson_comorbidity_index
")
```

```{r}
for (i in seq_len(nrow(dcolumns))) {
  x <- dcolumns[["medicalcoder"]][i]
  y <- dcolumns[["mimic"]][i]
  e <- base::substitute(identical(delta[[X]], delta[[Y]]), list(X = x, Y = y))
  print(e)
  r <- eval(e)
  print(r)
  stopifnot(r)
  delta[[x]] <- NULL
  delta[[y]] <- NULL
}
```

There are three comorbidities where there are different levels of severity.
`medicalcoder::comorbidities()` will set the less severe condition indicator to
0 when the more severe condition is flagged.  Both methods only consider the
more severe case in the index scoring.

Diabetes - `medicalcoder::comorbidities()` sets the flag for diabetes without
complication to 0 when diabetes with complication is present.  MIMIC code
retains the non-complex case.
```{r}
delta[dm == 1L & dmc == 1L, .N == 0L] # medicalcoder
delta[diabetes_without_cc == 0L & diabetes_with_cc == 0L, .N > 0L] # MIMIC

delta[diabetes_without_cc == 0L & diabetes_with_cc == 0L, .N > 0L & all(dm == 0L) & all(dmc == 0L)]
delta[diabetes_without_cc == 1L & diabetes_with_cc == 0L, .N > 0L & all(dm == 1L) & all(dmc == 0L)]
delta[diabetes_without_cc == 0L & diabetes_with_cc == 1L, .N > 0L &                all(dmc == 1L)]
delta[diabetes_without_cc == 1L & diabetes_with_cc == 1L, .N > 0L &                all(dmc == 1L)]

delta[dm == 0L & dmc == 0L, .N > 0L & all(diabetes_without_cc == 0L) & all(diabetes_with_cc == 0L)]
delta[dm == 1L & dmc == 0L, .N > 0L & all(diabetes_without_cc == 1L) & all(diabetes_with_cc == 0L)]
delta[dm == 0L & dmc == 1L, .N > 0L &                                 all(diabetes_with_cc == 1L)]
delta[dm == 1L & dmc == 1L, .N == 0L]

delta[, dm := NULL]
delta[, dmc := NULL]
delta[, diabetes_with_cc := NULL]
delta[, diabetes_without_cc := NULL]
```

Cancer - `medicalcoder::comorbidities()` sets malignant cancer (mal) to 0 when
metastatic solid tumor (mst) is present.  MIMIC retains both flags.
```{r}
delta[mal == 1L & mst == 1L, .N == 0L] # medicalcoder
delta[malignant_cancer == 1L & metastatic_solid_tumor == 1L, .N > 0L] # MIMIC

delta[malignant_cancer == 0L & metastatic_solid_tumor == 0L, .N > 0L & all(mal == 0L) & all(mst == 0L)]
delta[malignant_cancer == 1L & metastatic_solid_tumor == 0L, .N > 0L & all(mal == 1L) & all(mst == 0L)]
delta[malignant_cancer == 0L & metastatic_solid_tumor == 1L, .N > 0L &                 all(mst == 1L)]
delta[malignant_cancer == 1L & metastatic_solid_tumor == 1L, .N > 0L &                 all(mst == 1L)]

delta[mal == 0L & mst == 0L, .N > 0L & all(malignant_cancer == 0L) & all(metastatic_solid_tumor == 0L)]
delta[mal == 1L & mst == 0L, .N > 0L & all(malignant_cancer == 1L) & all(metastatic_solid_tumor == 0L)]
delta[mal == 0L & mst == 1L, .N > 0L &                              all(metastatic_solid_tumor == 1L)]
delta[mal == 1L & mst == 1L, .N == 0L]

delta[, mal := NULL]
delta[, mst := NULL]
delta[, malignant_cancer := NULL]
delta[, metastatic_solid_tumor := NULL]
```

Liver disease - `medicalcoder::comorbidities()` sets the flag for mild liver
disease (mld) to 0 when moderate/severe liver disease (msld) is flagged.  MIMIC
retains both flags.
```{r}
delta[mld == 1L & msld == 1L, .N == 0L] # medicalcoder
delta[mild_liver_disease == 1L & severe_liver_disease == 1L, .N > 0L] # MIMIC

delta[mild_liver_disease == 0L & severe_liver_disease == 0L, .N > 0L & all(mld == 0L) & all(msld == 0L)]
delta[mild_liver_disease == 1L & severe_liver_disease == 0L, .N > 0L & all(mld == 1L) & all(msld == 0L)]
delta[mild_liver_disease == 0L & severe_liver_disease == 1L, .N > 0L &                 all(msld == 1L)]
delta[mild_liver_disease == 1L & severe_liver_disease == 1L, .N > 0L &                 all(msld == 1L)]

delta[mld == 0L & msld == 0L, .N > 0L & all(mild_liver_disease == 0L) & all(severe_liver_disease == 0L)]
delta[mld == 1L & msld == 0L, .N > 0L & all(mild_liver_disease == 1L) & all(severe_liver_disease == 0L)]
delta[mld == 0L & msld == 1L, .N > 0L &                                all(severe_liver_disease == 1L)]
delta[mld == 1L & msld == 1L, .N == 0L]

delta[, mld := NULL]
delta[, msld := NULL]
delta[, mild_liver_disease := NULL]
delta[, severe_liver_disease := NULL]
```

All that is left in the `delta` `data.frame` are the id.vars and the
`num_cmrb`, and `cmrb_flag`.  These columns are from
`medicalcoder::comorbidities()` and report the number of comorbidities flagged
and indicator for any comorbidity.

```{r}
str(delta)
```

# References

<!-- ----------------------------------------------------------------------- -->
<!--                              End of File                                -->
<!-- ----------------------------------------------------------------------- -->
